---
title: "bioMetaLin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bioMetaLin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This vignette help to understand what bioMetaLine package does and what functions it contains
```{r setup}
library(bioMetaLin)
```


## Sample Data
sampleData() has one argument p, it is number of genes(column) you want to generate for data frames. Default p = 100
```{r}
datalist<-sampleData(p=100) 
#datalist$Data1 
#datalist$Data2
#datalist$Data3

## I included this datalist in the package. One can extract sampleData(p=100) using data function. 

#data(sample_data)
#sample_data$Data1
#sample_data$Data3
#sample_data$Data2
```

The p_vector() function generates a p-value vector for a dataframe. It determines the type of statistical test that should be applied to each biomarker based on the number of groups and type of data. 
The function takes two arguments: df, which is a dataframe, and stat_test, which is a Boolean indicating whether you want to display the type of test being applied. By default, stat_tests is set to TRUE

```{r}
p.vector<-p_vector(datalist[[1]],stat_test = T)
p.vector$p_value
p.vector$stat_test
```


the p_matrix() function generates a p-value matrix for a data list. For each dataframe, it determines the type of statistical test should be applied to each biomarket 
For example: 
- if dataframe is continuous and has more than 2 groups, it will choose ANOVA or Kruskal Wallis test based on assumptions.  
- if dataframe is continous and has 2 groups, it will choose t-test or Wilcoxon rank sum test based on assumptions.  
- if dataframe is categorical, it will choose Fisher's exact test or chi-squared test based on assumptions.
The function takes two arguments: datalist and stat_test, which is a Boolean indicating whether you want to display the type of test being applied. By default, stat_test is set to TRUE
```{r}
p.matrix<-p_matrix(datalist, stat_test = T)
p.matrix$p_value
p.matrix$stat_test
```
## Error and Warning Messages
All error and warning messages are contained within the p_matrix function, which is used by all of the statistical methods in this package. A data list must meet all of the criteria in p_matrix function first for all of the methods. 

## Statistical Function 
The following functions will be demonstrated using a sample data from the package.


### Fisher
Fisher's method is used to combine the results from several independence tests bearing upon the same overall hypothesis (H0). The method sums up the log-transformed p-values obtained from individual studies.
${X_{fisher}^2 = \sum\limits_{i=1}^{k}\ln(p_i)}$
fisher() take three arguments (datalist, stat = FALSE, p.adj=FALSE) 
```{r}
# no p-value adjustment
f<-fisher(datalist,stat=T)
f$statistic
f$p_value
```

```{r}
# bonferroni adjustment
f_bon<-fisher(datalist,stat=T,p.adj = 'bonferroni')
f_bon$p_value
```

```{r}
# fdr adjustment
f_fdr<-fisher(datalist,stat=T,p.adj = 'fdr')
f_fdr$p_value
```

### Stouffer
Stouffer's method is used to combine the results from several independence tests bearing upon the same overall hypothesis (H0). The method sums the inverse normal transformed p-values.
${Z_{Stouffer} = \sum\limits_{i=1}^{k} \frac{Z_{i}}{\sqrt{k}}}$
stouffer() take three arguments (datalist, stat = FALSE, p.adj=FALSE) 
```{r}
# no p-value adjustment
s<-stouffer(datalist,stat=T)
s$statistic
s$p_value
```

```{r}
# bonferroni adjustment
s_bon<-stouffer(datalist,stat=T,p.adj = 'bonferroni')
s_bon$p_value
```

```{r}
# fdr adjustment
s_fdr<-stouffer(datalist,stat=T,p.adj = 'fdr')
s_fdr$p_value
```


### Weighted Stouffer
weight is the square root of the sample size, wi
${Z_{weighted~Stouffer} = \frac{\sum_{i=1}^{k} w_i Z_i}{\sqrt{\sum_{i=1}^{k} w_i^2}}}$
weighted_stouffer() take three arguments (datalist, stat = FALSE, p.adj=FALSE) 
```{r}
# no p-value adjustment
ws<-weighted_stouffer(datalist,stat=T)
ws$statistic
ws$p_value
```

```{r}
# bonferroni adjustment
ws_bon<-weighted_stouffer(datalist,stat=T,p.adj = 'bonferroni')
ws_bon$p_value
```

```{r}
# fdr adjustment
ws_fdr<-weighted_stouffer(datalist,stat=T,p.adj = 'fdr')
ws_fdr$p_value
```

### Minimum p-vlaue
The minP method takes the minimum p-vlaue among the K studies as the test statistic.
It follows a beta distributionwith degrees of freedom  ${\alpha = 1}$ and ${\beta = k}$ under the null hypothesis.
minP() take three arguments (datalist, stat = FALSE, p.adj=FALSE) 
```{r}
# no p-value adjustment
Min<-minP(datalist,stat=T)
Min$statistic
Min$p_value
```

```{r}
# bonferroni adjustment
Min_bon<-minP(datalist,stat=T,p.adj = 'bonferroni')
Min_bon$p_value
```

```{r}
# fdr adjustment
Min_fdr<-minP(datalist,stat=T,p.adj = 'fdr')
Min_fdr$p_value
```


### Maximum p-value 
The maxP method takes maximum p-value as the test statistic.It follows a beta distribution with degrees of freedom ${\alpha = K}$ and ${\beta = 1}$ under the null hypothesis
maxP() take three arguments (datalist, stat = FALSE, p.adj=FALSE) 
```{r}
# no p-value adjustment
Max<-maxP(datalist,stat=T)
Max$statistic
Max$p_value
```

```{r}
# bonferroni adjustment
Max_bon<-maxP(datalist,stat=T,p.adj = 'bonferroni')
Max_bon$p_value
```

```{r}
# fdr adjustment
Max_fdr<-maxP(datalist,stat=T,p.adj = 'fdr')
Max_fdr$p_value
```

### Lancaster
Lancaster's method generalized Fisher's method by assigning different weights using the additive of chi-squared distribution.${T = \sum\limits_{i=1}^{k} [X_{n_i}^{2}]^{-1}(1-p_i)}$ where ${[X_{n_i}^{2}]^{-1}}$ is the inverse cumulative chi-square distribution function with $n_{i}$ degrees of freedom
lancaster() take three arguments (datalist, stat = FALSE, p.adj=FALSE) 
```{r}
# no p-value adjustment
lan<-lancaster(datalist,stat=T)
lan$statistic
lan$p_value
```

```{r}
# bonferroni adjustment
lan_bon<-fisher(datalist,stat=T,p.adj = 'bonferroni')
lan_bon$p_value
```

```{r}
# fdr adjustment
lan_fdr<-fisher(datalist,stat=T,p.adj = 'fdr')
lan_fdr$p_value
```


